Under-approximation of reachability through static analysis in multivalued asynchronous networks

### 2013-03-29 parts/def-ph.tex
minor fixes
remove most comments

% Process Hitting sans priorités
\begin{comment}
\begin{definition}[Process Hitting]\label{def:PH}
A \emph{Process Hitting} is a triple $(\PHs,\PHl,\PHa)$:
\begin{itemize}
\item $\PHs \DEF \{a,b,\dots\}$ is the finite set of \emph{sorts};
\item $\PHl \DEF \prod_{a\in\PHs} \PHl_a$ is the set of states with $\PHl_a = \{a_0,\dots,a_{l_a}\}$
the finite set of \emph{processes} of sort $a\in\Sigma$ and $l_a$ a positive integer with
	$a\neq b\Rightarrow \forall(a_i,b_j)\in\PHl_a\times\PHl_b,a_i\neq b_j$;
\item $\PHa \DEF \{ \PHfrappe{a_i}{b_j}{b_k}, \dots \mid
					(a,b)\in\PHs^2 \wedge (a_i,b_j,b_k)\in \PHl_a\times\PHl_b\times\PHl_b$ \\
	\hspace*{2cm} $\wedge b_j\neq b_k \wedge a=b\Rightarrow a_i=b_j\}$
			is the finite set of \emph{actions}.
\end{itemize}
$\PHproc$ denotes the set of all processes ($\PHproc \DEF \{ a_i\mid a\in\PHs \wedge a_i\in\PHl_a\}$).
\end{definition}
\end{comment}

@@

\begin{definition}[Process Hitting $n$-reduction] \towrite{Revoir terminologie : “reduction” ? “restriction” ?}
  If $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$ is a Process Hitting with $k$ priorities and $n \in \segm{1}{k}$, we denote $\restriction{\PH}{n}$ \towrite{$\PH_{|n}$ $\PH'$} the $n$-reduction of $\PH$ the following Process Hitting with $n$ priorities:
  $\restriction{\PH}{n} = (\PHs; \PHl; \restriction{\PHa}{n})$,
  with:
  $$\restriction{\PHa}{n} = (\PHa^{(1)}; \dots; \PHa^{(n)})$$
  Furthermore, we denote: $\restriction{\Sce}{n}$ the set of scenarios of actions of $\restriction{\PHa}{n}$.
%  
%  Furthermore, we define: $\restriction{\PH}{0} = (\PHs; \PHl; \restriction{\PHa}{0})$, where $\restriction{\PHa}{0} = ()$ is a $0$-uplet.
\end{definition}

\begin{comment} %example*
\pref{fig:runningPH-1} represents a PH $(\PHs,\PHl,\PHa)$ with
$\PHs = \{a,b,c\}$,
$\PHl_a = \{a_0,a_1,a_2\}$,
$\PHl_b = \{b_0, b_1\}$,
$\PHl_c = \{c_0, c_1\}$, and
\begin{align*}
\PHa & = \{
	\PHfrappe{a_2}{b_1}{b_0},
&&  \PHfrappe{b_0}{a_2}{a_1},
&&	\PHfrappe{c_0}{a_2}{a_1},\\
&&& \PHfrappe{b_0}{a_1}{a_0},
&&	\PHfrappe{c_0}{a_1}{a_0},\\
&&& \PHfrappe{b_1}{a_0}{a_1},
&&	\PHfrappe{c_1}{a_0}{a_1},\\
&&& \PHfrappe{b_1}{a_1}{a_2},
&&	\PHfrappe{c_1}{a_1}{a_2} \}\enspace.
\end{align*}
The action $h=\PHfrappe{b_1}{a_1}{a_2}$ is playable in the state
$s = \PHstate{b_1,a_1,c_0}$; and $s\PHplay h=\PHstate{b_1,a_2,c_0}$.
%
\begin{figure}[t]
\centering
\scalebox{1.3}{
\begin{tikzpicture}
\TSort{(0,0)}{a}{3}{r}
\TSort{(-3,0.5)}{b}{2}{l}
\TSort{(3,0.5)}{c}{2}{r}
%
\THit{b_1}{very thick}{a_0}{.west}{a_1}
\THit{b_1}{very thick}{a_1}{.north west}{a_2}
\THit{b_0}{}{a_2}{.west}{a_1}
\THit{b_0}{}{a_1}{.west}{a_0}
%
\path[bounce, bend left=60]
\TBounce{a_1}{very thick}{a_2}{.south}
\TBounce{a_0}{very thick}{a_1}{.south}
;
\path[bounce, bend right=60]
\TBounce{a_2}{}{a_1}{.north}
\TBounce{a_1}{}{a_0}{.north}
;
%
\THit{c_1}{very thick}{a_0}{.east}{a_1}
\THit{c_1}{very thick}{a_1}{.north east}{a_2}
\THit{c_0}{}{a_2}{.east}{a_1}
\THit{c_0}{}{a_1}{.east}{a_0}
%
\path[bounce, bend right=60]
\TBounce{a_1}{very thick}{a_2}{.south east}
\TBounce{a_0}{very thick}{a_1}{.south east}
;
\path[bounce, bend left=60]
\TBounce{a_2}{}{a_1}{.north}
\TBounce{a_1}{}{a_0}{.north east}
;
%
\THit{a_2}{bend right}{b_1}{.north east}{b_0}
\path[bounce, bend left=80]
\TBounce{b_1}{out=100,in=140}{b_0}{.north}
;
\end{tikzpicture}
}
\caption{\label{fig:runningPH-1}
A Process Hitting (PH) example.
Sorts are represented by labeled boxes, and processes by circles (ticks are
the identifiers of the processes within the sort, for instance, $a_0$ is the
process ticked $0$ in the box $a$).
An action (for instance $\PHfrappe{b_1}{a_1}{a_2}$) is represented by a pair of
directed arcs, having the hit part ($b_1$ to $a_1$) in plain line and the bounce
part ($a_1$ to $a_2$) in dotted line.
Actions involving $b_1$ or $c_1$ are in thick lines.
%The current state is represented by the grayed processes:
%$\PHstate{a_0,b_1,c_0,d_0}$.
}
\end{figure}

This PH example actually models a BRN where the component $a$ has three qualitative
levels and components $b$ and $c$ are boolean.
In this BRN, $b$ and $c$ activate $a$, while $a$ inhibits $b$.
The inhibition of $b$ by $a$ is only effective when $a$ is at level $2$;
in the other cases, $b$ cannot evolve in any direction.
The activation of $a$ by $b$ ($c$) is encoded by the actions making the level of $a$ increase (resp.
decrease) when $b$ ($c$) is present (resp. absent).
It is worth noticing that the activation of $a$ by $b$ ($c$) is independent from $c$ ($b$).
This may express a lack of knowledge on the cooperation between these two regulators:
we thus model an over-approximation of the possible actions.
\end{example*}
\end{comment}

@@

\begin{comment}
\begin{figure}[p]
\centering
\scalebox{1.3}{
\begin{tikzpicture}
\TSort{(0,0)}{b}{2}{t}
\TSort{(0,-3.8)}{c}{2}{b}
\TSort{(4.5,-3)}{a}{3}{r}

\TSetTick{bc}{0}{00}
\TSetTick{bc}{1}{01}
\TSetTick{bc}{2}{10}
\TSetTick{bc}{3}{11}
% \TSetSortLbcel{bc}{$\neg a\wedge b$}
\TSort{(-0.5,-2)}{bc}{4}{b}

\THit{b_1}{very thick,bend right}{bc_0}{.north}{bc_2}
\THit{b_1}{very thick,bend right}{bc_1}{.north}{bc_3}
\THit{b_0}{}{bc_2}{.north west}{bc_0}
\THit{b_0}{}{bc_3}{.north west}{bc_1}

\THit{c_0}{}{bc_1}{.south}{bc_0}
\THit{c_0}{}{bc_3}{.south}{bc_2}
\THit{c_1}{very thick}{bc_0}{.south}{bc_1}
\THit{c_1}{very thick}{bc_2}{.south}{bc_3}

\path[bounce, bend right=25]
\TBounce{bc_2}{}{bc_0}{.north east}
\TBounce{bc_3}{}{bc_1}{.north east}
;
\path[bounce, bend left=80, distance=30]
\TBounce{bc_0}{very thick}{bc_2}{.north}
\TBounce{bc_1}{very thick}{bc_3}{.north}
;
\path[bounce, bend right]
\TBounce{bc_0}{very thick}{bc_1}{.west}
\TBounce{bc_2}{very thick}{bc_3}{.west}
;
\path[bounce, bend left]
\TBounce{bc_3}{}{bc_2}{.east}
\TBounce{bc_1}{}{bc_0}{.east}
;

\THit{bc_3}{}{a_1}{.west}{a_2}
\path[bounce, bend left=40]
\TBounce{a_1}{}{a_2}{.south west}
;

\end{tikzpicture}
}

\caption{\label{fig:PH-cooperativity}
\towrite{Ajouter priorités}
A PH modeling a cooperativity between $b_1$ and $c_1$ to make
$a_1$ bounce to $a_2$.
Actions involving $b_1$ or $c_1$ are in thick lines.
}
\end{figure}
\end{comment}

\begin{comment}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \TSort{(0,0)}{a}{2}{l}
    \TSort{(2,0)}{b}{2}{r}
    \TSort{(4,0)}{c}{2}{r}

    \THit{a_1}{prio}{b_1}{.west}{b_0}
    \THit{b_1}{}{c_0}{.north west}{c_1}

    \path[bounce, bend right]
    \TBounce{b_1}{}{b_0}{.north west}
    ;
    \path[bounce, bend left]
    \TBounce{c_0}{}{c_1}{.south west}
    ;
  \end{tikzpicture}
  \label{fig:ph-conca}
  \caption{A simple PH example depicting “backward concurrency” between two actions belonging to different classes of priorities.}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \TSort{(0,0)}{a}{2}{l}
    \TSort{(2,0)}{b}{2}{r}
    \TSort{(4,0)}{c}{3}{r}

    \THit{a_1}{}{b_0}{.west}{b_1}
    \THit{b_0}{prio}{c_1}{.south west}{c_0}
    \THit{b_1}{prio}{c_1}{.west}{c_2}

    \path[bounce, bend left]
    \TBounce{b_0}{}{b_1}{.south west}
    \TBounce{c_1}{}{c_2}{.south west}
    ;
    \path[bounce, bend right]
    \TBounce{c_1}{}{c_0}{.north west}
    ;
  \end{tikzpicture}
  \label{fig:ph-concb}
  \caption{A simple PH example depicting “forward concurrency” between two actions belonging to different classes of priorities.}
\end{figure}
\end{comment}

\begin{comment}
\begin{figure}[p]
\centering
\scalebox{1.3}{
\begin{tikzpicture}
\path[use as bounding box] (-4,-1.9) rectangle (4.5,3.9);
%
\TSort{(0,0)}{a}{3}{l}
\TSort{(3, 3)}{b}{2}{t}
\TSort{(3,-1)}{c}{2}{b}
%
\TSetTick{bc}{0}{00}
\TSetTick{bc}{1}{01}
\TSetTick{bc}{2}{10}
\TSetTick{bc}{3}{11}
% \TSetSortLbcel{bc}{$\neg a\wedge b$}
\TSort{(-3,-0.5)}{bc}{4}{l}
%
\THit{bc_3}{}{a_1}{.north west}{a_2}
\THit{bc_0}{}{a_1}{.south west}{a_0}
\path[bounce]
\TBounce{a_1}{bend left}{a_2}{.south west}
\TBounce{a_1}{bend right}{a_0}{.north west}
;
%
\THit{b_0}{}{a_2}{.east}{a_1}
\THit{b_1}{}{a_0}{.north east}{a_1}
\path[bounce]
\TBounce{a_2}{bend left}{a_1}{.north east}
\TBounce{a_0}{bend right=20}{a_1}{.south}
;
%
\THit{c_0}{bend right}{a_2}{.south east}{a_1}
\THit{c_1}{bend right}{a_0}{.east}{a_1}
\path[bounce]
\TBounce{a_2}{bend left=20}{a_1}{.north}
\TBounce{a_0}{bend right=30}{a_1}{.south east}
;
%
\path[dashed,hit]
	(2,-1.3) edge[bend left=10] (-2.3,-0.7)
	(2.2, 3.3) edge[bend right=10] (-2.3,3)
;
%
\THit{a_2}{bend left,out=40,in=80}{b_1}{.north west}{b_0}
\path[bounce, bend right]
\TBounce{b_1}{}{b_0}{.east}
;
%
\end{tikzpicture}
}
%
\caption{\label{fig:runningPH-2}
PH resulting from the refinement of the one in \pref{fig:runningPH-1} by the
specification of several cooperations.
The actions from $b$ and $c$ to the cooperative sort $bc$ are identical to those defined in
\pref{fig:PH-cooperativity} and are represented here by a single dashed arc.
}
\end{figure}
\end{comment}

@@

%In this section, we describe several restrictions.
%Although the definition of a Process Hitting with $k$ priorities allows a wide range of models, the method given in the next section only applies to Process Hitting models that regard these restrictions.

\begin{comment}
We define the set of chains of actions $\PHh^n(b, a)$ between two sorts $b$ and $a$, with $n \in \segm{1}{k}$, as below:
\begin{align*}
\PHh^n(b, a) = \{ (h_i)_{i \in \segm{0}{s}} \in (\restriction{\PHh}{n})^{s+1} &\mid s \in \sN, \PHsort(\PHhitter(h_0)) = b \wedge \PHsort(\PHtarget(h_s)) = a \\
  & \wedge \forall i \in \segm{1}{s}, \PHsort(\PHtarget(h_{i-1})) = \PHsort(\PHhitter(h_{i})) \}
\end{align*}

We define the neighbor \towrite{Revoir terminologie : “neighborhood” ? “vicinity” ?} sorts of any sort $a$ and $n \in \segm{1}{k}$ as :
$$
\Vs^n(a) = \{ b \in \PHs \mid \PHh^m(b, a) \neq \emptyset \wedge m = \min(n, \prio(a)) \} \cup \{ a \}
$$

And, as a shortcut:
$$
\Vs(a) = \Vs^{\prio(a)}(a)
$$

We define the neighbor actions of any sort $a$ as :
$$
\Vh(a) = \{ h \in \restriction{\PHh}{\prio(a)} \mid \{ \PHsort(\PHhitter(h)), \PHsort(\PHtarget(h)) \} \subset \Vs(a) \wedge \prio(h) \leq \prio(a) \}
$$
\end{comment}

%\paragraph{Cycle-freeness of $\restriction{\PH}{k-1}$}
%\towrite{Définition formelle ? + justification}
%The model $\restriction{\PH}{k-1}$ contains no cycles. Cycles are still allowed in $\PH$ however.

In the scope of this paper, we focus on a specific class of Process Hitting models.
We define here the restrictions that lead to this class of models \todo{and show that they are equivalent to discrete networks}.
%Let $\PH$ be a PH model on which we want to apply the static analysis.
%In this paper, we focus on a specific class of models that are equivalent to boolean networks.

@@

We denote $\pfp_s(a)$ the set of possible local steady-states of a sort $a$ from a state $s$.
\todo{À corriger : état global ? + prendre en compte l'état de la sorte}
\begin{align*}
  \pfp_s(a) = \{ a_i \in \PHl_a &\mid \exists \delta \in \restriction{\Sce}{1}, \PHget{(s \PHplay \delta)}{a} = a_i \\
  & \quad \wedge \forall \PHhit{b_i}{c_j}{c_k} \in \Vh(a), \PHget{(s \PHplay \delta)}{b} \neq b_i \vee \PHget{(s \PHplay \delta)}{c} \neq c_j \}
\end{align*}
Of course, if $a \in \components$, then $\pfp_s(a) = \PHget{s}{a}$.

@@

\begin{comment}
\paragraph{Bounded termination}
The dynamics of model $\restriction{\PH}{k-1}$ contains no cycles. Cycles are still allowed in $\PH$ however.
Formally : $\exists N \in \sN, \forall \delta \in \restriction{\Sce}{k-1}, |\indexes{\delta}| \leq N$.

\paragraph{Local priorities}
\towrite{Justification}
All actions hitting the same sort belong to the same class of priority:
$$\forall h_1, h_2 \in \PHh, \PHtarget(h_1) = \PHtarget(h_2) \Rightarrow \prio(h_1) = \prio(h_2)$$
Furthermore, for all sort $a$, we denote $\prio(a)$ the priority of the actions hitting $a$, if any:
$$\forall a \in \PHs, \exists h \in \PHh, \PHtarget(h) = a \Rightarrow \prio(a) = \prio(h)$$
\towrite{Ou : $\nexists h \Rightarrow \prio(a) = 0$}

\paragraph{Local stable states}
\towrite{Justification}
We define $\pfp$ as the set of possible stable states of a sort from a given state:
\begin{align*}
  \pfp_\ctx = \{ s \PHplay \delta \in \PHl &\mid s \in \ctx \wedge \delta \in \restriction{\Sce}{k-1} \\
  & \qquad \wedge \forall \PHhit{b_i}{c_j}{c_k} \in \restriction{\PHh}{k-1}, \PHget{(s \PHplay \delta)}{b} \neq b_i \vee \PHget{(s \PHplay \delta)}{c} \neq c_j \}
  %\text{$h$ n'est pas jouable dans $s \PHplay \delta$} \}
\end{align*}

All processes are a stable state of a sub-state of the predecessors:
$\forall a_i \in \PHproc, \prio(a) < k \Rightarrow \exists s \in \restriction{\PHl}{\Vs(a)}, a_i \in \pfp_s$.
We note: $\stable_{a_i} = \{ s \in \restriction{\PHl}{\Vs(a)} \mid a_i \in \pfp_s \}$.
\end{comment}

\begin{comment}
These restrictions bring new interesting properties to the PH models regarding them.

\pref{th:vplay} tells that for any action $h = \PHhit{a_i}{b_j}{b_k}$, if no action can be played in the neighborhood of $a$ in a given state, then $h$ can be played after a series of hits that do not prevent it to be fired. Furthermore, if the requisites of \pref{th:vplay} are true, then no action of this series of hits belongs to $V(a)$, as stated in \pref{co:vplay}.
\begin{theorem}
\label{th:vplay}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, (\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j \wedge $\\
  $\forall h' \in \Vh^{\prio(b)-1}(a), \PHhitter(h') \notin s \wedge \PHtarget(h') \notin s) \Rightarrow$\\
  $(\exists \delta \in \Sce, \text{$h$ can be played in $s \PHplay \delta$})$
\end{theorem}

\begin{proof}
  Because of the cycle-freeness of $\restriction{\PH}{k-1}$, $\restriction{\PH}{\prio(a)}$ is also cycle-free and there is a $\delta \in \Sce$ so that $\forall h' \in \PHh$, $h'$ cannot be played in $s \PHplay \delta$. Furthermore, $\forall h' \in \delta, \PHsort(\PHtarget(h')) \neq a$ by construction of $\Vs(a)$.
\end{proof}

\begin{corollary}
  \label{co:vplay}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, (\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j \wedge $\\
  $\forall h' \in \Vh^{\prio(b)-1}(a), \PHhitter(h') \notin s \wedge \PHtarget(h') \notin s) \Rightarrow$\\
  $\exists \delta \in \Sce, (\forall n \in \indexes{\delta}, \delta_n \notin \Vh(a)) \wedge (\text{$h$ can be played in $s \PHplay \delta$})$
\end{corollary}

\pref{th:totalss} states that from any state, a steady state is eventually reached. A local variant of this theorem, given in \pref{co:totalss}, can be derived,
where, for all $\Omega \subset \PHs$, $$\restriction{\PHl}{\Omega} = \underset{a \in \Omega}{\times} \PHl_a$$ and
for all $H \subset \PHh$, $$\restriction{\Sce}{H} = \{ \delta \in \Sce \mid \forall n \in \indexes{\delta}, \delta_n \in H \} \enspace .$$
\begin{theorem}
\label{th:totalss}
  $\forall n \in \segm{1}{k-1}, \forall s \in \PHl, \exists \delta \in \restriction{\Sce}{n}$, no action of model $\restriction{\PH}{n}$ can be played in $s \PHplay \delta$.
\end{theorem}

\begin{corollary}
\label{co:totalss}
  $\forall a \in \PHs, \forall s \in \restriction{\PHl}{\Vs(a)}, \exists \delta \in \restriction{\Sce}{\Vh(a)}$, no action of $\Vh(a)$ can be played in $s \PHplay \delta$.
\end{corollary}

\pref{th:autohits} states that any sequence $\zeta$ of self-actions can be played eventually.
\begin{theorem}
\label{th:autohits}
  Let $a \in \PHs$, $s \in \PHl$ and $\zeta \in \BS(a)$, with $n = |\indexes{\zeta}|$ so that $\forall i \in \segm{1}{n}, \PHsort(\PHhitter(\zeta_i)) = a$.
  %If no action can be played in $V(a)$ in state $s$, then
  $\exists (\delta_i)_{i \in \segm{1}{n}} \in \Sce$ so that $\forall i \in \segm{1}{n}$, $\zeta_i$ can be played in $s \PHplay \delta_1 \PHplay \zeta_1 \PHplay \dots \PHplay \delta_i$. 
\end{theorem}

\begin{proof}
  With \pref{co:vplay} applied iteratively.
\end{proof}

\towrite{Théorème nécessaire pour la suite: une séquence d'actions dans $\BS(a)$ peut être jouée si précédée ou entrelacée par des $\zeta$ (?)}
\end{comment}

>>>>

@@

\begin{definition}[Process Hitting $n$-reduction] \towrite{Revoir terminologie : “reduction” ? “restriction” ?}
  If $\PH = (\PHs; \PHl; \PHa^{\langle k \rangle})$ is a Process Hitting with $k$ priorities and $n \in \segm{1}{k}$, we denote $\restriction{\PH}{n}$ \towrite{$\PH_{|n}$ $\PH'$} the $n$-reduction of $\PH$ the following Process Hitting with $n$ priorities:
  $\restriction{\PH}{n} = (\PHs; \PHl; \restriction{\PHa}{n})$,
  with:
  $$\restriction{\PHa}{n} = (\PHa^{(1)}; \dots; \PHa^{(n)})$$
  Furthermore, we denote: $\restriction{\Sce}{n}$ the set of scenarios of actions of $\restriction{\PHa}{n}$.
\end{definition}

@@

@@

In the scope of this paper, we focus on a specific class of Process Hitting models.
We define here the restrictions that lead to this class of models \todo{and show that they are equivalent to discrete networks}.

@@

We denote $\pfp_s(a)$ the set of possible local steady-states of a sort $a$ from a state $s$.
\todo{À corriger : état global ?}
\begin{align*}
  \pfp_s(a) = \{ a_i \in \PHl_a &\mid \exists \delta \in \restriction{\Sce}{1}, \PHget{(s \PHplay \delta)}{a} = a_i \\
  & \quad \wedge \forall \PHhit{b_i}{c_j}{c_k} \in \Vh(a), \PHget{(s \PHplay \delta)}{b} \neq b_i \vee \PHget{(s \PHplay \delta)}{c} \neq c_j \}
\end{align*}
Of course, if $a \in \components$, then $\pfp_s(a) = \PHget{s}{a}$.

@@
